---
title: "Bird Sightings At Sea"
description: ""
author: "Steven Wolf"
date: "4/14/2026"
categories:
  - TidyTuesday
  - R 
execute: 
  messages: false
  warning: false
  echo: false
---


I'm back doing some [#TidyTuesday](https://www.tidytuesday.com/) visualizations. This week let's get bird-brained. [The dataset](https://github.com/rfordatascience/tidytuesday/blob/main/data/2026/2026-04-14/readme.md) has 2 objects (I have already joined the wind and sea data to the ships data):

- `birds` Observation data related to bird sightings. Includes things like bird count, plumage of birds, behavior of birds.
- `ships` Observation data related to the location of the ship, date and time of observation, weather conditions, and sea conditions.

```{r}
load('Tongarewa.RData')
library(tidyverse)
library(maps)
library(viridis)
library(GGally)
library(showtext)
library(ggtext)

showtext_opts(dpi = 225, regular.wt = 300, bold.wt = 800)

theme_set(theme_classic(base_size = 14, base_family = 'sans'))

font_add("fa7-brands", 
         "/usr/share/fonts/fontawesome-free-7.0.1-desktop/otfs/Font Awesome 7 Brands-Regular-400.otf")
showtext_auto(enable = TRUE)

gh <- str_glue("<span style='font-family: \"fa7-brands\"'>&#xf09b;</span>")
li <- str_glue("<span style='font-family:\"fa7-brands\"'>&#xf08c;</span>")
bluesky <- str_glue("<span style='font-family:\"fa7-brands\"'>&#xe671;</span>")
ttText <- str_glue("**#TidyTuesday** &bull; Source: {bluesky} David Hood")
caption_text <- str_glue("{ttText} <br> {gh} datawolf04 {li} wolf-253b6625a")
ttStart <- str_glue("<b>#TidyTuesday</b>")

theme_mine = function(...){
  theme_minimal() %+replace%
    
    theme(
      legend.title = element_markdown(size = rel(0.8), hjust = 0.5),
      legend.text = element_text(size = rel(0.7)), 
          
      plot.margin = margin(t = 20, r = 20, b = 20, l = 20),
      plot.title = element_text(size = rel(1.25),
                                margin = margin(0,0,10,0),hjust=0.5),
      plot.subtitle = element_text(size = rel(1.25),
                                margin = margin(0,0,10,0),hjust=0.5),
      axis.title.x = element_text(margin = margin(10, 0, 0, 0), size = rel(0.8)),
      axis.title.y = element_text(margin = margin(0, 10, 0, 0), size = rel(0.8),angle=90),
      axis.text = element_text(size=rel(0.8)),
      plot.caption = element_markdown(
        size        = rel(.7),
        family      = "sans",
        color       = "grey25",
        lineheight  = 0.6,
        hjust       = 1,
        halign      = 0,
        margin      = margin(t = 10, b = 5)
      )
    ) 
}


```

## Snapshot of birds data
This dataset is full of missing data. Indeed if I were to drop any row that doesn't have an value in every column, that analysis would be quite boring:

```{r}
birds |> drop_na()
```

Let's look at what the worst offenders are:
```{r}
empty <- birds |> summarise(across(everything(), ~ mean(is.na(.x)))) |> t() |> data.frame() |> 
    rownames_to_column("observation")
names(empty)[2] = 'naFrac'
empty |> arrange(desc(naFrac)) |> filter(naFrac > 0.2)
```
That's essentially everything, except the count.

```{r}
empty |> arrange(desc(naFrac)) |> filter(naFrac <= 0.2)
```

## Summary of ship data
I will begin by getting a sense of when and where this data is being taken.

```{r}
world_map <- map_data("world", wrap=c(-180,180))

ggplot() +
    geom_polygon(data = world_map, aes(x = long, y = lat, group = group), 
        fill = "lightgray", color = "white") +
    geom_point(data = ships, aes(x=longitude, y=latitude, color = date)) +
    labs(title= "Observation Locations", color="Observation Date",x = "", y="") + 
    scale_color_viridis_c(trans = 'date') + 
    coord_fixed(
        xlim = c(min(ships$longitude, na.rm = TRUE)-15, 180),
        ylim = c(-90,10),
        ratio = 1.2
    ) +
        theme_mine() +
    theme(axis.title = element_blank(), 
          axis.text = element_blank(),
          legend.position = "inside",
          legend.position.inside = c(1.05,1),
          legend.justification = c(1,1))
```

## Seasonal patterns in bird observations
Since I've been playing with the map functions, I decided to plot count data on the map for different seasons in the dataset.

```{r}
combined <- left_join(birds, ships, by="record_id") |> filter(!is.na(season)) |> 
    mutate(season = str_to_sentence(season))

pltTitle <- str_glue('Bird count data from {format(min(combined$date),"%B %Y")} to {format(max(combined$date),"%B %Y")}')

ggplot() +
    geom_polygon(data = world_map, aes(x = long, y = lat, group = group), 
        fill = "lightgray", color = "white") +
    geom_point(data = combined, aes(x=longitude, y=latitude, color = count), na.rm=TRUE) +
    labs(title= pltTitle, color="Number of birds", x='', y='',
         caption=caption_text) + 
    scale_color_viridis_c(option="turbo", trans="log10") + 
    coord_fixed(
        xlim = c(min(ships$longitude, na.rm = TRUE)-15, 180),
        ylim = c(-90,10),
        ratio = 1.2
    ) +
    facet_wrap(vars(season), nrow = 2, ncol=2) +
    theme_mine() + 
    theme(axis.title = element_blank(), 
          axis.text = element_blank(),
    ) 
ggsave('birdObsBySeason.png')
```



